---
title: "ST 558 Homework 3"
format:
  html:
    toc: true
    theme: cosmo
editor_options: 
  chunk_output_type: console
---

## Task 1: Conceptual Questions

1.  If your working directory is `myfolder/homework/`, what relative path would you specify to get the file located at `myfolder/MyData.csv`?

- The relative path would be `"../MyData.csv"`. The `..` tells R to move up one level out of the current folder `homework/` and into the parent folder `myfolder/`.

2.  What are the major benefits of using R projects?

- R Projects improve organization, reproducibility, collaboration, and file management

3.  What is Git and what is GitHub?

- **Git** is a version control system that tracks changes to files over time. It allows you to:

  - Save snapshots of your work (commits)

  - View the history of changes

  - Revert to previous versions if something goes wrong

  - Work on different features independently using branches

  - Merge changes from multiple contributors

- **GitHub** is a cloud-based platform that hosts Git repositories online. It allows you to:

  - Store and back up your Git repositories

  - Collaborate with others on projects

  - Share code publicly or privately

  - Review changes through pull requests

  - Track issues and manage projects

4.  What are the two main differences between a `tibble` and a `data.frame`?

- A tibble is a modern version of a data frame that displays data more cleanly by only showing the first few rows and columns, along with each column's data type. Tibbles also preserve their structure when subsetting, whereas data frames may simplify the result to a vector when a single column is selected.

5.  Rewrite the following nested function call using baseR's chaining operator:

```{r}
library(tidyverse)
arrange(filter(select(as_tibble(iris), starts_with("Petal"), Species), Petal.Length < 1.55), Species)
```

```{r}
iris |>
  as_tibble() |>
  select(starts_with("Petal"), Species) |>
  filter(Petal.Length < 1.55) |>
  arrange(Species)
```

6.  What is meant by long format data and wide format data? Which do we generally prefer for statistical analysis?

- Long format data has multiple rows for a single record, with one column holding variable names and another holding the values. Wide format data has all variables and repeated measurements for a single entity spread across a single row. For most statistical analyses, we generally prefer long format data because it is easier to work with and is compatible with many statistical modeling and visualization functions.

# Task 2 Reading Delimited Data

Read the delimited data into R using `read_delim`

```{r}
library(tidyverse)
glass_data <- read_delim("https://www4.stat.ncsu.edu/~online/datasets/glass.data", 
                       delim = ",", col_names = c("Id", "RI", "Na", "Mg", "Al", "Si", "K", "Ca", "Ba", "Fe", "Type_of_Glass"))

glass_data
```

Create a factor version of the variable that replaces the numbers with string labels

```{r}
glass_data |>
  mutate(Type_of_Glass = factor(Type_of_Glass,
                                levels = (1:7),
                                labels = c("building_windows_float_processed",
                                           "building_windows_non_float_processed",
                                           "vehicle_windows_float_processed",
                                           "vehicle_windows_non_float_processed",
                                           "containers",
                                           "tableware",
                                           "headlamps")))
```

Keep only observations where the `Fe` variable is less than 0.2 and the `Type_of_Glass` is either “tableware” or “headlamp”

```{r}
glass_data |>
  mutate(Type_of_Glass = factor(Type_of_Glass,
                                levels = (1:7),
                                labels = c("building_windows_float_processed",
                                           "building_windows_non_float_processed",
                                           "vehicle_windows_float_processed",
                                           "vehicle_windows_non_float_processed",
                                           "containers",
                                           "tableware",
                                           "headlamps"))) |>
  filter(Fe < 0.2 & Type_of_Glass %in% c("tableware", "headlamp"))
```

Read in the fixed width file containing yeast data

```{r}
yeast_data <- read_fwf("https://www4.stat.ncsu.edu/online/datasets/yeast.data",
                       fwf_widths(c(12, 6, 6, 6, 6, 6, 6, 6, 6, 3), 
                                  c('seq_name', 'mcg', 'gvh', 'alm', 'mit', 'erl', 'pox', 'vac', 'nuc',                                           'class')))
yeast_data
```

select all columns apart from `seq_name` and `nuc`

```{r}
yeast_data |>
  select(-seq_name, -nuc)
```

Find the mean and median, grouped by `class`, of all the numeric variables in the dataset

```{r}
yeast_data |>
  select(-seq_name, -nuc) |>
  group_by(class) |>
  mutate(across(where(is.numeric), list(mean = mean, median = median), .names = "{.col}_{.fn}")) 
```

# Task 3: Combining Excel and Delimited Data

Read in the white wine data

```{r}
library(readxl)
white_wine_data <- read_excel("white-wine.xlsx")
white_wine_data
```

Read in the variable names from the 2nd sheet of the same file. Apply those names to the `white_wine_data`

```{r}
column_names_white_wine_data <- read_excel("white-wine.xlsx", 
                                     sheet = excel_sheets("white-wine.xlsx")[2])
column_names_white_wine_data

colnames(white_wine_data) <- column_names_white_wine_data$Variables
white_wine_data
```

Add a column to the dataset specifying the type of wine

```{r}
white_wine_data$wine_type <- "white"
View(white_wine_data)
```

Read in the red wine data

```{r}
library(readr)
red_wine_data <- read_delim("red-wine.csv", delim = ';')
red_wine_data
```

Apply the names from the second sheet of the `white_wine_data` to the `red_wine_data`

```{r}
colnames(red_wine_data) <- column_names_white_wine_data$Variables
red_wine_data
```

Add a column to the dataset specifying the type of wine

```{r}
red_wine_data$wine_type <- "red"
View(red_wine_data)
```

Join the two wine datasets

```{r}
wine_data <- dplyr::bind_rows(white_wine_data, red_wine_data)
wine_data
```

Filter out records with `quality > 6.5` or `alcohol < 132`

```{r}
wine_data |>
  filter(quality > 6.5, alcohol < 132)
```

Arrange the data in descending order on `alcohol`

```{r}
wine_data |>
  filter(quality > 6.5, alcohol < 132) |>
  arrange(desc(quality))
```

Select specific variables to include 

```{r}
wine_data |>
  filter(quality > 6.5, alcohol < 132) |>
  arrange(desc(quality)) |>
  select(contains("acid"), alcohol, wine_type, quality)
```

find the mean and sd for alchohol, grouped by `quality`

```{r}
wine_data |>
  filter(quality > 6.5, alcohol < 132) |>
  arrange(desc(quality)) |>
  select(contains("acid"), alcohol, wine_type, quality) |>
  group_by(quality) |>
  mutate(across(alcohol, list(mean = mean, sd = sd), .names = "{.col}_{.fn}"))
```


# Task 4: Database Practice

Connect to the lahman database

```{r}
library(DBI)
library(tidyverse)

con <- dbConnect(RSQLite::SQLite(), "lahman.db")
dbListTables(con)
```

Return all of the data from the Teams table for the year 2015

```{r}
teams <- tbl(con, "Teams")

teams |>
  filter(yearID == 2015)
```

Do it in SQL now

```{r}
tbl(con,
    sql("SELECT * FROM Teams WHERE yearID = 2015"))
```

Return all of the players in the hall of fame, the year they were voted into the hall of fame, and their
category 

```{r}
hof <- tbl(con, "HallOfFame") |>
  filter(inducted == "Y") |>
  select(playerID, yearID, category) |>
  collect()

hof
```

Combine the table from the previous question with the People table in order to have the nameFirst
and nameLast names added to what was returned above

```{r}
people <- tbl(con, "People") |>
            select(playerID, nameFirst, nameLast) |>
            collect()

hof <- left_join(hof, people) |>
       select(nameFirst, nameLast, playerID, yearID, category)

hof
```

Return only the playerID, G, W, and L columns from the Managers table. Use chaining to complete some other tasks, including determining the overall win/loss records for every hall of fame manager

```{r}
mgrs <- tbl(con, "Managers") |>
        select(playerID, G, W, L) |>
        group_by(playerID) |>
        summarize(G_managed = sum(G, na.rm = TRUE),
                  Total_W = sum(W, na.rm = TRUE),
                  Total_L = sum(L, na.rm = TRUE)) |>
        collect() |>
        mutate(career_wl = Total_W/G_managed) |>
        arrange(desc(career_wl))

mgrs
```

Return information for only the people that managed a team and were inducted into the hall of fame

```{r}
hof_mgrs <- inner_join(mgrs, hof) |>
            select(playerID, nameFirst, nameLast, yearID,
                   category, G_managed, Total_W, Total_L, career_wl)

# close the db connection
dbDisconnect(con)

hof_mgrs
```

