---
title: "ST 558 Homework 3"
format:
  html:
    toc: true
    theme: cosmo
---

## Chris's Version

## Task 1: Conceptual Questions

1. You should specify the path: `"../MyData.csv"`
2. Using R projects increases the **"portability"** of your work by keeping necessary files in a common directory.
3. Git is a **version control protocol**, GitHub is a **service that hosts Git repositories**
4. Tibbles **print nicer** than data frames and do not **simplify into a vector** when a single column is referenced
5. Here's the same code with chaining:
```{r}
library(tidyverse)

as_tibble(iris) |>
  select(starts_with("Petal"), Species) |>
  filter(Petal.Length < 1.55) |>
  arrange(Species) |>
  head()
```
6. Long format data contains a single observation per row, while wide format data contains multiple observations within the same row. We prefer long format data because it's easier to work with.

## Task 2: Reading Delimited Data

```{r}
glass <- read_csv("https://www4.stat.ncsu.edu/online/datasets/glass.data",
                  col.names = c("id", "ri", "na", "mg", "al", "si", "k", "ca", "ba", "fe", "type")) |>
         as_tibble()


```




