---
title: "ST 558 Homework 3"
format:
  html:
    toc: true
    theme: cosmo
---

## Task 1: Conceptual Questions

1. You should specify the path: `"../MyData.csv"`. The `..` tells R to move up one level out of the current folder `homework/` and into the parent folder `myfolder/`.
2. Using R projects increases the **"portability"** of your work by keeping necessary files in a common directory for use with version control software or sharing with collaborators
3. **Git** is a version control system that tracks changes to files over time. It allows you to:

- Save snapshots of your work (commits)
- View the history of changes
- Revert to previous versions if something goes wrong
- Work on different features independently using branches
- Merge changes from multiple contributors

**GitHub** is a cloud-based platform that hosts Git repositories online. It allows you to:

- Store and back up your Git repositories
- Collaborate with others on projects
- Share code publicly or privately
- Review changes through pull requests
- Track issues and manage projects
4. Tibbles:
   1) **Print nicer** than data frames
   2) Do not **simplify to a vector** when a single column is referenced
5. Here's the same code with chaining:
```{r}
# shut up tidyverse
suppressMessages(library(tidyverse))

as_tibble(iris) |>
  select(starts_with("Petal"), Species) |>
  filter(Petal.Length < 1.55) |>
  arrange(Species) |>
  head()
```
6. Long format data has multiple rows for a single record, with one column holding variable names and another holding the values. Wide format data has all variables and repeated measurements for a single observational unit spread across a single row. For most statistical analyses, we generally prefer long format data because it is easier to work with and is compatible with many statistical modeling and visualization functions.

## Task 2: Reading Delimited Data

### Glass Data

We'll start by reading in the `glass.csv` file:
```{r}
glass_names <- c("id", "ri", "na", "mg", "al", "si", "k", "ca", "ba", "fe", "type")

glass <- read_csv("https://www4.stat.ncsu.edu/online/datasets/glass.data",
                  col_names = glass_names,
                  show_col_types = FALSE)

glass
```

Our tibble looks good, but we need to recode our `type` variable.

I wanted to use a switch statement to recode the levels, but those don't work inside of `mutate` for some reason. I found `case_match` in the `dplyr` documentation as a vectorized alternative so we're using it here.
```{r}
glass |>
  mutate(glass, type = factor(type,
                              levels = 1:7,
                              labels = c("building_windows_float_processed",
                                         "building_windows_non_float_processed",
                                         "vehicle_windows_float_processed",
                                         "vehicle_windows_non_float_processed",
                                         "containers",
                                         "tableware",
                                         "headlamps"))) |>
                filter(fe < 0.2, type %in% c("tableware", "headlamps"))
```

### Yeast Data

Start by reading in the fixed-width data as shown in the video:
```{r}
yeast_names <- c("seq_name", "meg", "gvh", "alm", "mit", "erl", "pox", "vac", "nuc", "class")

yeast <- read_fwf("https://www4.stat.ncsu.edu/online/datasets/yeast.data",
                  col_positions = fwf_widths(c(12, 6, 6, 6, 6, 6, 6, 6, 6, 3),
                                             col_names = yeast_names),
                  show_col_types = FALSE)

yeast
```

Now we can subset the columns and create the summaries. I'm barely changing the code from the video:
```{r}
yeast_sub <- yeast |>
              select(-c("seq_name", "nuc")) |>
              group_by(class) |>
              mutate(across(where(is.numeric), list(mean = mean, median = median),
                            .names = "{.col}_{.fn}"))

yeast_sub
```

Great, we have our summary columns for each variable grouped by the class.

## Task 3: Joining Data

We start by importing the wine data from the first sheet of the excel workbook we're provided:
```{r}
library(readxl)

white_wine <- read_xlsx("white-wine.xlsx",
                        sheet = 1)

white_wine
```

We don't want variable names with spaces in them. Let's pull better names from the second sheet and replace these weird ones.

```{r}
wine_names <- read_xlsx("white-wine.xlsx",
                        sheet = 2) |>
              pull()

colnames(white_wine) <- wine_names

white_wine
```

Great! We just need to add the type of wine to each observation:

```{r}
white_wine$type <- "White"

# preview
white_wine[,10:13]
```

Now that the white wine data is complete, we can turn our attention to the reds:
```{r}
red_wine <- read_delim("https://www4.stat.ncsu.edu/~online/datasets/red-wine.csv",
                       skip = 1,
                       delim = ";",
                       col_names = wine_names)

red_wine$type <- "red"

red_wine
```

Perfect! With the correct column types and names, it's easy to merge our two data sets:

```{r}
wine <- bind_rows(white_wine, red_wine)

wine
```

Lastly, we'll make a pipeline using the combined data set:
```{r}
wine |>
  filter(quality > 6.5, alcohol < 132) |> # only 6.5+ quality, less than 132 alcohol
  arrange(desc(quality)) |>                         # arrange in desc quality order
  select(contains("acid"), alcohol, type, quality) |>      # subset columns
  group_by(quality) |>                                    # group by quality
  mutate(across(alcohol, list(mean = mean, sd = sd),    # add alcohol mean/sd columns
                .names = "{.fn}_alcohol"))
```

## Task 4: Database Practice

Connecting to the `lahman` database is quite easy:
```{r}
library(DBI)

lahman <- dbConnect(RSQLite::SQLite(), "lahman.db")

dbListTables(lahman)
```
There are a lot of tables within this database!

We start by querying the `Teams` table and returning observations from the year 2015:
```{r}
data_2015 <- tbl(lahman, "Teams") |>
               filter(yearID == 2015) |>
               collect()

data_2015
```

We can do the same thing using SQL commands instead:

```{r}
data_2015_frmsql <- tbl(lahman, sql(
                        "SELECT *
                           FROM `Teams`
                           WHERE (`yearID` = 2015)"
                        )) |>
                    collect()

data_2015_frmsql
```
This is the same table from above, so it looks like I still remember how to write SQL! I'm going to use the dplyr workflow going forward because I'm more comfortable with it.

### Hall of Fame Players
Here we're taking a look at the `HallOfFame` table and doing some wrangling:
```{r}
hof <- tbl(lahman, "HallOfFame") |>
         filter(inducted == "Y") |>
         select(playerID, yearID, category) |>
         collect()

hof
```

Next, we use the `playerID` key variable to left join the `hof` and `people` data sets so we have the first and last name for each player.
```{r}
people <- tbl(lahman, "People") |>
            select(playerID, nameFirst, nameLast) |>
            collect()

hof <- left_join(hof, people) |>
       select(nameFirst, nameLast, playerID, yearID, category)

hof
```

### Manager Statistics

We're now focusing our attention on the `Managers` table:
```{r}
mgrs <- tbl(lahman, "Managers") |>
        select(playerID, G, W, L) |>
        group_by(playerID) |>
        summarize(G_managed = sum(G, na.rm = TRUE),
                  Total_W = sum(W, na.rm = TRUE),
                  Total_L = sum(L, na.rm = TRUE)) |>
        collect() |>
        mutate(career_wl = Total_W/G_managed) |>
        arrange(desc(career_wl))

mgrs
```

Unsurprisingly, the highest career win/loss stats are held by managers who managed a few games to victory.

Lastly, we join the `hof` and `mgrs` tables by `playerID` to return all Hall of Fame members who managed games at some point in their career. I'm also reordering the columns so the order is more useful.
```{r}
hof_mgrs <- inner_join(mgrs, hof) |>
            select(playerID, nameFirst, nameLast, yearID,
                   category, G_managed, Total_W, Total_L, career_wl)

# be a good egg and close the db connection
dbDisconnect(lahman)

hof_mgrs
```

Great! This table contains every person inducted into the Hall of Fame who managed at least one game.